pirouette: the error BEAST2 makes in inferring a phylogeny

### 2019-02-25 article.tex
Updates from Overleaf

The development of new powerful inference tools, 
such as BEAST [\cite{drummond2007beast}] or MrBayes [\cite{huelsenbeck2001mrbayes}], 
provides us the possibility to build phylogenetic trees 
from genetic material extracted from extant organisms.
This has constituted an important step forward 
in our understanding on how species 
evolve.
Such tools have been increasingly exploited to hypothesize and test what are the main drivers
and modes for diversification processes.

There are plenty of diversification models: 
constant-birth [Yule, 19..],
birth-death [\cite{nee1994reconstructed}], 
time-dependent [Rabosky D., LASER package?], 
diversity dependent [\cite{etienne2011diversity}], 
protracted birth-death [\cite{rosindell2010protracted}][\cite{etienne2012prolonging}],
multiple-birth death [Laudanno, Bilderbeek, Haegeman \& Etienne, in preparation],
a rate shifting birth-death model [the SLS model. Laudanno, Haegeman \& Etienne, in preparation]
punctuated equilibrium (as known as "early burst") [\cite{harmon2010early}],
dependent on a binary trait [\cite{maddison2007estimating}], 
multiple state trait [\cite{fitzjohn2012diversitree}],
concealed state [\cite{beaulieu2016detecting}] 
or even a combination of multiple concealed and observable states [\cite{herrera2018detecting}].

Such models usually rely on the assumption 
that a particular diversification process 
can be explained mainly by a specific mechanism.
Such a mechanism is built into a likelihood function, 
which provides the likelihood of the parameters, given the data.
By maximizing it, it is possible 
to estimate the what are the best values for model parameters. 
It is a standard practice in assessing such models,
to verify the level of accuracy they can achieve
on simulated phylogenies, of which full knowledge is given.

Diversification models are getting more complex over time. There are many
reasons for this. One reason is the increase in computing power,
which allows this heightened complexity possible.
Another reason is to explore a process that are not yet considered by currently existing diversification models. An example of this
is the multiple-birth model [Laudanno et al., in preparation], 
that is the first model to allow simultaneous speciations to co-occur.

However, whenever a new model is proposed there is always a trade-off between the cost of increased complexity versus the benefits that it can provide.
Therefore desirable models have to be able to disentangle the intrinsic complexity of a system, 
highlight the main reasons of its behaviour and avoid unnecessary complexity. Furthermore developing models can cost time, human and computational resources.

@@

For this reason we believe it is important to have techniques to quantitatively assess the advancement provided by new models, in comparison to the existing ones.

@@

\verb;pirouette; is an R package that performs such an experiment,
and is built on \verb;babette; [Bilderbeek \& Etienne, 2018], 
which calls the popular Bayesian inference tool 
BEAST2 [\cite{bouckaert2014beast}]. With \verb;pirouette;, one
can easily measure the error made by Bayesian inference in recovering
any given phylogeny, helping us to validate new diversification models.

@@

The goal of \verb;pirouette; is to measure the inference error BEAST2
makes from a given reconstructed phylogeny. For this measurement,
two models need to be specified: a generative and an inference model.
A generative model is a combination of site model, clock model and tree prior,
used to simulate alignments. An inference model is a combination of 
site model, clock model, tree prior, an optional node calibration prior
and setup of the Markov chain Monte Carlo (MCMC) algorithm, 
used to produce a posterior. Node calibration priors, called 'most
recent common ancestor' (MRCA) priors within BEAST2 allow to create a 
time-calibrated phylogeny, by specifying a normal distribution around
an expected time when two or more taxa's ancestors diversified.

The pipeline is summarized by the following steps, which will be described in detail below:
\begin{enumerate}
    \item from a given phylogeny an alignment is simulated;
    \item from the alignment an inference model is chosen;
    \item the alignment is used as BEAST2 input to infer a posterior distribution of phylogenies;
    \item posterior phylogenies are compared with the given original phylogeny to estimate the error we make. The bigger the match, the lower the error;
\end{enumerate}
\richel{figure to pipeline here}
There is also the option to generate a 'twin tree',
that goes through the same pipeline.

@@

The second step selects an inference model, using the alignment.
The user can specify whether to use the generative model or select a best model
from a set of inference models. 
When selecting the generative model,
the site and clock model used in the alignment simulation are used
in the inference. Because the given phylogeny (on which the alignment is based)
may have followed any tree prior (i.e. speciation model), the user needs
to specify which tree prior is used in the inference. 
When selecting for the best
model, the alignment is used to find the inference model that has the
highest evidence (i.e. marginal likelihood) from a set of candidate inference models.
The evidence of an inference model is estimated using a nested sampling (NS)
approach, as described in \cite{maturana2018model}. The nested sampling is
performed by \verb;mcbette; \cite{mcbette}, that calls the 'NS' BEAST2 package. 
Using BEAST2 packages (in a scripted way) can only be done under Linux and Mac; 
inference model selection is not available on Windows.

@@

The fourth step measures the inference error, using the phylogenies in the
Bayesian posterior. These phylogenies are compared to the given
original phylogeny using an error statistic, which is the nLTT 
statistic (\cite{janzen2015approximate}) by default,
but also a custom error statistic is supplied, based
on the absolute difference in the gamma statistic [Pybus \& Harvy]. 
Additionally, the user can specify the
proportion of posterior phylogenies to 
discard (i.e. the burn-in), throwing away the first $10\%$
of all phylogenies by default. This burn-in is used to discard
the part of the MCMC chain that has not yet converged to a
representative part of state space.

@@

\begin{lstlisting}[
    language=R, floatplacement=H, 
    frame=single
    label = {lst:create_alignment}, 
    caption = {Create an alignment}
  ]
alignment_params <- create_alignment_params(
  root_sequence = create_blocked_dna(length = 1000),
  mutation_rate = 0.1
)
\end{lstlisting}

By default, an alignment is created using a Jukes-Cantor (JC) site model
and a strict clock model. A JC site model assumes that mutations
between all nucleotides are constant at an equal rate. 
A strict clock model assumes that
the mutation rate of all species is equal and constant.
Due to this, we state that the generative model for the alignment is
a JC site model and a strict clock model.

@@

\begin{table}
  \begin{tabular}{ | c | c | c | l | }
    \hline
    \textbf{model type} & \textbf{run if} & \textbf{measure evidence} & \textbf{inference model} \\ 
    \hline
    generative & always & no & JC, strict, Yule \\
    \hline
  \end{tabular}
  \caption{
    Experimental setup to answer the first research question.
    JC: Jukes-Cantor site model.
    strict: strict clock model.
    Yule: Yule (pure-birth) tree prior
  }
\end{table}

@@

\begin{lstlisting}[
  language=R, 
  floatplacement=H, frame=single,
  label = {lst:create_generative_experiment},
  caption = {
    Create a default experiment. This default experiment is
    for a generative model, that is always run, without its evidence measured,
    assuming a JC, strict, Yule inference model
  }
]
experiments <- list(create_experiment())
\end{lstlisting}

@@

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_2.png}
  \caption{
    The error BEAST2 makes from a phylogeny 
    when the generative and inference model are the same
  }
  \label{fig:example_1}
\end{figure}

@@

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
experiment <- create_all_experiments()
\end{lstlisting}

Also here, the third (the BEAST2 inference) and fourth (measuring the error)
step have sensible defaults, and we are not
interested in using a twin tree. We can create the complete
\verb;pirouette; parameter set (again) like this:

@@

The following settings are the same as in the previous section:
phylogeny (listing \ref{lst:unknown_phylogeny}), 
alignment parameters (listing \ref{lst:create_alignment}), 
experiments (listing \ref{lst:create_generative_experiment})
BEAST2 inference and error measuring parameters.

@@

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_4.png}
  \caption{
    The error BEAST2 makes from a phylogeny 
    picking the best inference model, compared to the background noise?
    Here, the twin column shows the error BEAST2 makes on an idealized
    tree, to measure the noise, which is the minimal error. 
  }
\end{figure}

@@

There are many other examples that can be found in the \verb;pirouette; 
documentation.

@@

\giovanni{We need to discuss the results we got from the 3 experiments.}
\richel{Only iff the journal requires this}

>>>>

The development of new powerful inference tools, 
such as BEAST [\cite{drummond2007beast}] or MrBayes [\cite{huelsenbeck2001mrbayes}], 
allowed to build phylogenetic trees 
from genetic material extracted from extant organisms.
This has constituted an important step forward 
in our understanding on how species 
evolve.
Such tools have been increasingly exploited to hypothesize and test what are the main drivers and modes for diversification processes.

There are plenty of diversification models: 
constant-birth [Yule, 19..],
birth-death [\cite{nee1994reconstructed}], 
time-dependent [Rabosky D., LASER package?], 
diversity dependent [\cite{etienne2011diversity}], 
protracted birth-death [\cite{rosindell2010protracted}][\cite{etienne2012prolonging}],
multiple-birth death [Laudanno, Bilderbeek, Haegeman \& Etienne, in preparation],
a rate shifting birth-death model [the SLS model. Laudanno, Haegeman \& Etienne, in preparation]
punctuated equilibrium (as known as "early burst") [\cite{harmon2010early}],
dependent on a binary trait [\cite{maddison2007estimating}], 
multiple state trait [\cite{fitzjohn2012diversitree}],
concealed state [\cite{beaulieu2016detecting}] 
or even a combination of multiple concealed and observable states [\cite{herrera2018detecting}]. \giovanni{we might need to polish this list a little bit}

Such models usually rely on the assumption 
that a particular diversification process 
can be explained mainly by a specific mechanism.
Such a mechanism is built into a likelihood function, 
which provides the probability to realize a process given the parameters.
By maximizing it, it is possible 
to estimate what are the best values for model parameters. 
It is a standard practice in assessing such models,
to verify the level of accuracy they can achieve
on simulated phylogenies, of which full knowledge is given.

Diversification models are getting more complex over time. This is facilitated by the increase in computing power due to the advancement of technology.
Increased complexity made possible the exploration of new models encompassing mechanisms that are not yet considered by previous diversification models. An example of this is the multiple-birth model [Laudanno et al., in preparation],
that is the first model to allow simultaneous speciations to co-occur.

However, whenever a new model is proposed there is always a trade-off between the cost of increased complexity versus the benefits that it can provide.
In fact developing models can cost time, human and computational resources.
Therefore desirable models have to be able to disentangle the intrinsic complexity of a system, highlight the main reasons of its behaviour and avoid unnecessary complexity.

@@

For this reason we believe it is important to have techniques to quantitatively assess the advancement provided by the introduction of new models, in comparison to the existing ones.

@@

\verb;pirouette; is an R package that performs such an experiment,
and is built on \verb;babette; [Bilderbeek \& Etienne, 2018], 
which calls the popular Bayesian inference tool 
BEAST2 [\cite{bouckaert2014beast}]. With \verb;pirouette;, one
can easily measure the error made by Bayesian inference in recovering
any given phylogeny, helping us to validate new diversification models.
\giovanni{I think we should mention the fact that BEAST2 allows to implement customized tree priors more complicated than the simple yule and bd. These tree priors, in the end, are what we call likelihood in our models, so it's pivotal in proposing our argument, I think.}

@@

The goal of \verb;pirouette; is to measure the inference error BEAST2
makes from a given reconstructed phylogeny \giovanni{We should mention here that the phylogenies that are meant to be submitted to pirouette are simulated phylogenies generated according to the model we want to test}. For this measurement, two models need to be specified: a generative and an inference model. 
A generative model is a combination of site model, clock model and tree prior,
used to simulate alignments. An inference model is a combination of 
site model, clock model, tree prior, an optional node calibration prior
and setup of the Markov chain Monte Carlo (MCMC) algorithm, 
used to produce a posterior. Node calibration priors, called 'most
recent common ancestor' (MRCA) priors within BEAST2 allow to create a 
time-calibrated phylogeny, by specifying a normal distribution around
an expected time when two or more taxa's \giovanni{I have the feeling that taxa should be plural here. Should we mention that the group must be monophyletic?} ancestors diversified.

\subsection{pirouette's pipeline}
The pipeline is summarized by the following steps, which will be described in detail below:
\begin{enumerate}
    \item from a given phylogeny an alignment is simulated;
    \item from the alignment an inference model is chosen;
    \item the alignment is used as BEAST2 input to infer a posterior distribution of phylogenies;
    \item posterior phylogenies are compared with the given original phylogeny to estimate the error we make. The higher the match, the lower the error;
\end{enumerate}
\richel{figure to pipeline here}
There is also the option to generate a 'twin tree',
that goes through the same pipeline.

@@

The second step selects an inference model, using the alignment.
The user can specify whether to use the generative model or select the best model
from a set of inference models. 
When selecting the generative model,
the site and clock model used in the alignment simulation are used
in the inference. Because the given phylogeny (on which the alignment is based)
may have followed any tree prior (i.e. speciation model), the user needs
to specify which tree prior is used in the inference. 
When selecting for the best
model, the alignment is used to find the inference model that has the
highest evidence (i.e. marginal likelihood) from a set of candidate inference models.
The evidence of an inference model is estimated using a nested sampling (NS)
approach, as described in \cite{maturana2018model}. The nested sampling is
performed by \verb;mcbette; [\cite{mcbette}], that calls the 'NS' BEAST2 package. 
Using BEAST2 packages (in a scripted way) can only be done under Linux and Mac; 
inference model selection is not available on Windows.

@@

The fourth step measures the inference error, using the phylogenies in the
Bayesian posterior. These phylogenies are compared to the given
original phylogeny using an error statistic, which is the nLTT 
statistic (\cite{janzen2015approximate}) by default,
but also a custom error statistic is supplied, based
on the absolute difference in the gamma statistic [\cite{pybus2000testing}]. 
Additionally, the user can specify the
proportion of posterior phylogenies to 
discard (i.e. the burn-in), throwing away the first $10\%$
of all phylogenies by default. This burn-in is used to discard
the part of the MCMC chain that has not yet converged to a
representative part of state space.

@@

\begin{lstlisting}[
    language=R,
    floatplacement=H, frame=single,
    label = {lst:create_alignment}, 
    caption = {Create an alignment}
  ]
alignment_params <- create_alignment_params(
  root_sequence = create_blocked_dna(length = 1000),
  mutation_rate = 0.1
)
\end{lstlisting}

By default, an alignment is created using a Jukes-Cantor (JC) site model
and a strict clock model. A JC site model assumes that mutation rates from any nucleotide to any other are equal and constant. A strict clock model assumes that the mutation rates of all species are equal and constant.
Due to this, we state that the generative model for the alignment is
a JC site model and a strict clock model.

@@

\begin{table}
  \begin{tabular}{ | c | c | c | l | }
    \hline
    \textbf{model type} & \textbf{run if} & \textbf{measure evidence} & \textbf{inference model} \\ 
    \hline
    generative & always & no & JC, strict, Yule \\
    \hline
  \end{tabular}
  \caption{
    Experimental setup to answer the first research question.
    JC: Jukes-Cantor site model.
    strict: strict clock model.
    Yule: Yule (pure-birth) tree prior.
  }
\end{table}

@@

\begin{lstlisting}[
  language=R, 
  floatplacement=H, frame=single,
  label = {lst:create_generative_experiment},
  caption = {
    Create a default experiment. This default experiment is
    for a generative model, that is always run, without its evidence measured,
    assuming a JC site model, strict clock model, Yule inference model.
  }
]
experiments <- list(create_experiment())
\end{lstlisting}

@@

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_2.png}
  \caption{
    The error BEAST2 makes from a phylogeny 
    when the generative and inference model are the same.
  }
  \label{fig:example_1}
\end{figure}

@@

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
experiments <- create_all_experiments()
\end{lstlisting}

Also here, the third (the BEAST2 inference) and fourth (measuring the error)
steps have sensible defaults, and we are not
interested in using a twin tree. We can create the complete
\verb;pirouette; parameter set (again) like this:

@@

The following settings are the same as in the previous section:
phylogeny (listing \ref{lst:unknown_phylogeny}), 
alignment parameters (listing \ref{lst:create_alignment}), 
experiments (listing \ref{lst:create_generative_experiment}),
BEAST2 inference and error measuring parameters.

@@

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_4.png}
  \caption{
    The error BEAST2 makes from a phylogeny 
    picking the best inference model, compared to the background noise.
    Here, the twin column shows the error BEAST2 makes on an idealized
    tree, to measure the noise, which is the minimal error. 
  }
\end{figure}

@@

There are many other examples that can be found in the \verb;pirouette; 
documentation.
\giovanni{it could be nice to add a listing with simple commands that refer to the most useful examples}

@@

\giovanni{We need to discuss the results we got from the 3 experiments.}
\richel{Only if the journal requires this}
